### 2022-12-07 paper/writeup.tex
algorithm section added

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{tikz}
\usetikzlibrary{arrows, snakes,backgrounds}
\tikzstyle{place}=[circle,draw=black,thick, inner sep=0pt, minimum size = 5mm]

@@

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}

@@

\section{Notation}
    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
     $N$ and $E$ are the sets of nodes and edges respectively,
     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome. 
    Additionally, a node $x\in N$ is \textit{unary} if it only has one child, 
    ie. only a single edge extending below it on the tree.
    With these tree sequences our goal is to
    minimize the number of edges
    as well as infer coalescence of genes. 
    To remove edges we consider the following example.
    Suppose for a tree $T_k$ in the tree sequence $TS$ 
    there exists a branch which contains a unary node $b\in N$.
    Then the node $b$ has one parent, $a\in N$, 
    and one child $c\in N$,
    each with edges $e_{ab}, e_{bc}\in E$ connecting them.
    If there is an edge between the nodes $a$ and $c$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $e_{ab},e_{bc}$ 
    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
    This action reduces the length of the edge $e_{ac}$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such unary nodes 
    over the entire tree sequence.

\begin{tikzpicture} %% Do this with phylogenetic trees using tikz (WILL NEED LUALATEX COMPILER) %%
\node[place] (a)  {a};
\node[place] (b)  [below of = a] {b};
\node[place] (c)  [below of = b] {c};
\draw [-] (a.south) -- (b.north);
\draw [-] (b.south) -- (c.north);
\end{tikzpicture}

    In full detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$
    be a tree sequence.

\begin{algorithm}  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??
\caption{Extend Edges}\label{alg:cap}
\begin{algorithmic}
\SetKwFunction{mirror}{Mirror Coordinates}
\State   $ N' \gets N$
\State   $ e'\in E' \gets E $
\State   $ T'_k \gets T_{n-k} $
\State   $ T' \gets (\{T'_k\}_{k=1}^n,N',E',L) $ 
\end{algorithmic}
\end{algorithm}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usetikzlibrary{arrows, snakes,backgrounds}
\tikzstyle{place}=[circle,draw=black,thick, inner sep=0pt, minimum size = 5mm]

@@

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}

@@

\subsection*{Notation}
    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
     $N$ and $E$ are the sets of nodes and edges respectively,
     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome.
     The length of each tree $T_k$ is some subinterval $[a_k,a_{k+1})$ 
     with the lengths of consecutive trees in the sequence 
     only intersecting each intervals endpoints. 
    Additionally, a node $x\in N$ is \textit{unary} if it only has one child, 
    ie. only a single edge extending below it on the tree.
    With these tree sequences our goal is to
    minimize the number of edges
    as well as infer coalescence of genes. 
    To remove edges we consider the following example.
    Suppose for a tree $T_k$ in the tree sequence $TS$ 
    there exists a branch which contains a unary node $b\in N$.
    Then the node $b$ has one parent, $a\in N$, 
    and one child $c\in N$,
    each with edges $e_{ab}, e_{bc}\in E$ connecting them.
    If there is an edge between the nodes $a$ and $c$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $e_{ab},e_{bc}$ 
    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
    This action reduces the length of the edge $e_{ac}$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such unary nodes 
    over the entire tree sequence.
\begin{center}
\begin{tikzpicture} 
	\node[circle,draw,inner sep=1pt] (a) at (0,2) {a}
	[growth parent anchor=south] 
		child { node[circle,draw,inner sep=1pt] {b} 		
	 			child {
	 				node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
	 					child{node (d) at (0,1) {$T_k$}
	 						edge from parent [draw=none]}
	 				edge from parent 
	 					node[left] {$e_{b,c}$}
	 			}	 		
	 		edge from parent
	 			node[left] {$e_{a,b}$}
	 	};

	\node[circle,draw,inner sep=1pt] (a') at (3,2) {a}
	[growth parent anchor=south]
		child {
			child{
				node[circle,draw,inner sep=1pt] (c') at (0,-0.25) {c}	
					child{node (d) at (0,1) {$T_{k+1}$}
						edge from parent [draw=none]}
			}
			node[left] {$e_{a,c}$}
		};

	\node at (6,0) {$\Longrightarrow$};
	\node at (6,0.5) {\tn{Edge Extend}};

	\node[circle,draw,inner sep=1pt] (a''') at (9,2) {a}
	[growth parent anchor=south] 
	child { node[circle,draw,inner sep=1pt]	 {b}	
		child {
			node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
			child{node (d) at (0,1) {$T_k$}
				edge from parent [draw=none]}
			edge from parent 
			node[left] {$e_{b,c}$}
		}
		edge from parent
		node[left] {$e_{a,b}$}
	};

	\node[circle,draw,inner sep=1pt] (a'v) at (12,2) {a}
	[growth parent anchor=south] 
	child { node[circle,draw,inner sep=1pt] {b}		
		child {
			node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
			child{node (d) at (0,1) {$T_{k+1}$}
				edge from parent [draw=none]}
			edge from parent 
			node[left] {$e_{b,c}$}
		}
		edge from parent
		node[left] {$e_{a,b}$}
	};

\end{tikzpicture}
\end{center}
    In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
    be a tree sequence.
    Within tskit we can track a tree sequence using a function called `EdgeDiffs'.
    This function tracks edges which are removed and added 
    between consecutive trees in the sequence. 
    As in \ref{} (tikzfigure), we can track if 
    $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 
    when creating $T_{k+1}$, 
    and we can see if $e_{a,c}$ was just added in $T_{k+1}$.

\begin{algorithm}[!ht]  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??

\SetStartEndCondition{ }{}{}%
\SetKwProg{Fn}{def}{\string:}{}
\SetKwFunction{Range}{range}%%
\SetKw{KwTo}{in}
\SetKwFor{For}{for}{\string:}{}%
\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
\SetKwFor{While}{while}{:}{fintq}%
\SetKw{Break}{break}
\newcommand\forcond{$i$ \KwTo\Range{$n$}}
\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%

\caption{Extends edges across a tree sequence $TS$.}\label{alg:edge}
\SetKwFunction{ex}{Extend Edges}
\SetKwFunction{fwd}{Forward Extend}
\DontPrintSemicolon
\LinesNotNumbered
%\SetAlgoNoEnd
\input{A tree sequence $TS$}
\output{New tree sequence,$TS'$, with reduced number of edges}

\BlankLine

\Fn{\fwd (tree sequence $TS$, forwards = \tn{True})}{
{	
\eIf{forwards=\tn{True}}{EdgeDiffs $\gets TS$.edgediffs()\BlankLine}
{\BlankLine
	EdgeDiffs $\gets$ reversed($TS$.edgediffs())\;}
}
\Indm \For{(Interval, EdgesOut, EdgesIn) \KwTo EdgeDiffs}{
	\For{e1 \KwTo EdgesOut}{
		\For{e2 \KwTo EdgesOut}{
			\If{e2.parent = e1.child}{
				\For{e3 \KwTo EdgesIn}{
					\If{e3.parent=e1.parent \tn{AND} e3.child=e2.child}{
						e1.right $\gets$ Interval.right\;
						e2.right $\gets$ Interval.right\;
						e3.left $\gets$ Interval.right\;
						\If{e3.left = e3.right}{
							$E$.remove(e3)\;
							}}}}}}}\;
$TS'\gets TS$.update()\;						
\KwRet{$TS'$}\;
}
\BlankLine						
\Fn{\ex ($TS$, iter=100)}{
	edges $\gets$ $TS$.numedges()\;
	\BlankLine
	\For{$*$ \KwTo \Range(iter)}{
		$TS' \gets TS$.\fwd(fowards=\tn{True})\;
		$TS' \gets TS$.\fwd(fowards=\tn{False})\;
		\eIf{$TS$.numedges = edges}{
			\Break\BlankLine}
		{edges $\gets$ $TS'$.numedges()\;}
		}
}
\KwRet{$TS'$}\;
\end{algorithm}
